---
title: "To Err is Human: Decoding Type I and Type II Error"
subtitle: "Unveiling the Truth Behind Statistical Decisions"
format: 
  revealjs:
    slide-number: true
    smaller: true
    chalkboard: true
    scrollable: true
editor: source
---



## Red Marble Proportion Estimation


$$H_0: p = 0.5$$
$$H_a: p \neq 0.5$$

```{r}
#| echo: false
#| message: false
library(tidyverse)
source("../utils.R")

red_marble = "🔴"
blue_marble = "🔵"

prob_red = 0.64
num_marbles = 5000

set.seed(42)
marbles = sample(c(red_marble, blue_marble), size = num_marbles, 
                 replace = TRUE, prob = c(prob_red, 1 - prob_red)) 
```



```{r}
#| echo: true

set.seed(7595)
n = 100
one_sample = sample(marbles, n)
p_hat = mean(one_sample == red_marble)
print(str_glue("Percentage of red marbles: {p_hat * 100}%"))

z_score = (p_hat - 0.5) / sqrt(0.5 * (1 - 0.5) / n)
print(str_glue("z-score: {round(z_score, 3)}"))

p_val = (1 - pnorm(p_hat, mean = 0.5, sd = 0.05)) * 2
print(str_glue("p-value: {round(p_val, 3)}"))
```


Since $-1.96 < 1.6 < 1.96$ or $0.11 > 0.05$, we fail to reject the null hypothesis at the 0.05 level 
and conclude that the proportion of red marbles is 50%.

## Type I and II Errors

| | $H_0$ is true | $H_0$ is false |
| --- | --- | --- |
| Reject $H_0$ | Type I error | Correct decision |
| Fail to reject $H_0$ | Correct decision | Type II error |


## Practice (1)
:::{.incremental}
- $H_0$ : the beverage is fine to drink
- $H_a$ : the beverage is too hot to drink
- **Type I error**: You think the beverage is too hot so you decide to wait, but by the time you drink it, it's too cold.
- **Type II error**: You decide the beverage is fine to drink now, but it's too hot and you burn your tongue.
:::

## Practice (2)
:::{.incremental}
- $H_0$: the new drug has no effect on the disease
- $H_a$: the new drug has effect on the disease
- **Type I error**: the results of the experiment wrongly suggest that the drug is effective.
- **Type II error**: the results shows no effect on the disease while the drug is effective.
:::


## Practice (3)
:::{.incremental}
- $H_0$: The patient does not have the condition.
- $H_a$: The patient has the condition.
- **Type I error**: A healthy patient is diagnosed with the condition (False Positive).
- **Type II error**: A sick patient is diagnosed as healthy (False Negative).
:::


## Type I Error Probability

$$\text{Prob}(\text{Reject } H_0 \mid H_0 \text{ is true}) = \alpha$$

```{r}
#| echo: true
critical_region_plot(0.5, 0.05)
```





## Type II Error Probability

$$\text{Prob}(\text{Do not reject } H_0 \mid \mu = \hat p)$$

```{r}
#| echo: false
p = critical_region_plot(0.5, 0.05) + 
  geom_vline(xintercept = 0.5, color = "black", linetype="dashed")

p1 = p + 
  geom_point(x = p_hat, y = 0, color = "blue", size = 3)
```


## Type II Error Probability


```{r}
#| echo: true

se = sqrt(p_hat * (1 - p_hat) / n)
p1 +
  stat_function(fun = dnorm, args = list(mean = p_hat, sd = se), color = "blue") +
  geom_vline(xintercept = p_hat, color = "blue", linetype = "dashed") +
  geom_point(x = p_hat, y = 0, color = "blue", size = 3)
  

```



## Type II Error Probability

```{r}
#| echo: true

ls = 0.5 - 1.96 * 0.05
rs = 0.5 + 1.96 * 0.05

p2=p1 + stat_function(fun = dnorm, args = list(mean = p_hat, sd = se), color = "blue") +
  stat_function(fun = function(x) dnorm(x, p_hat, se), 
                xlim = c(ls, rs), fill = "gray50", geom = "area") +
  geom_vline(xintercept = p_hat, color = "blue", linetype = "dashed") +
  geom_point(x = p_hat, y = 0, color = "blue", size = 3) + 
  annotate("text", x = 0.545, y = 2, label = "Type II Error", color = "white")
p2
```

. . .

```{r}
#| echo: true

prob = pnorm(rs, mean = p_hat, sd = se) - pnorm(ls, mean = p_hat, sd = se)
print(str_glue("Prob(Type II Error) = {round(prob, 3)}"))

```

## Statistical Power

The Power of a test is the probability of rejecting a false null hypothesis.

$$\text{Power} = \text{Prob}(\text{Reject } H_0 \mid \mu = \hat p) = 1 - \text{Prob(Type II Error)}$$


```{r}
#| echo: true
power = 1 - prob
print(str_glue("Power = {round(power, 3)}"))
```

## Another Example


```{r}
#| echo: true

p_hat = 0.69; se = sqrt(p_hat * (1 - p_hat) / n)
p3 = p + stat_function(fun = dnorm, args = list(mean = p_hat, sd = se), color = "blue") +
  stat_function(fun = function(x) dnorm(x, p_hat, se), 
                xlim = c(ls, rs), fill = "gray50", geom = "area") +
  geom_vline(xintercept = 0.69, color = "blue", linetype = "dashed") +
  geom_point(x = 0.69, y = 0, color = "blue", size = 3)
p3
```



```{r}
#| echo: true

prob = pnorm(rs, mean = p_hat, sd = se) - pnorm(ls, mean = p_hat, sd = se)
power = 1 - prob
print(str_glue("Prob(Type II Error) = {round(prob, 3)}; Power = {round(power, 3)}"))
```


## Power and Sample Size

$$\bar X \sim \mathcal{N}(\mu, \frac{\sigma^2}{n})$$

```{r}
#| echo: true

n = 100
se = sqrt(0.5 * (1 - 0.5) / n)
print(str_glue("Standard Error (n = {n}) = {round(se, 3)}"))

n = 400
se = sqrt(0.5 * (1 - 0.5) / n)
print(str_glue("Standard Error (n = {n}) = {round(se, 3)}"))

```



## Power and Sample Size


```{r}
#| echo: true

n = 400
p_hat = 0.58
se = sqrt(p_hat * (1 - p_hat) / n)

ls = 0.5 - 1.96 * sqrt(0.5 * (1 - 0.5) / n)
rs = 0.5 + 1.96 * sqrt(0.5 * (1 - 0.5) / n)

p4 = critical_region_plot(0.5, sqrt(0.5 * (1 - 0.5) / n)) + 
  geom_vline(xintercept = 0.5, color = "black", linetype="dashed") +
  geom_point(x = 0.58, y = 0, color = "blue", size = 3) + 
  stat_function(fun = dnorm, args = list(mean = p_hat, sd = se), color = "blue") +
  stat_function(fun = function(x) dnorm(x, p_hat, se), 
                xlim = c(ls, rs), fill = "gray50", geom = "area") +
  geom_vline(xintercept = p_hat, color = "blue", linetype = "dashed") +
  geom_point(x = p_hat, y = 0, color = "blue", size = 3) + 
  annotate("text", x = 0.535, y = 0.5, label = "Type II Error", color = "white")
p4
```

## Power and Sample Size

```{r}
#| echo: true

prob = pnorm(rs, mean = p_hat, sd = se) - pnorm(ls, mean = p_hat, sd = se)
power = 1 - prob
print(str_glue("Prob(Type II Error) = {round(prob, 3)}; Power = {round(power, 3)}"))
```



## Trade-off between Type I and II Errors

- $\alpha$ is set before the experiment.
- Typical values for $\alpha$ are 0.05 and 0.01.

## Trade-off between Type I and II Errors

- Decreasing $\alpha$ increases $\beta$.

```{r}
p4
```



## Trade-off between Type I and II Errors

- Increasing sample size decreases $\beta$.
- Typical values for $\beta$ are 0.2 and 0.1.

```{r}
grid.arrange(p2, p4, ncol=2)
```



