---
title: "Unmasking the Truth: Diving into Hypothesis Testing"
subtitle: "How to verify your friend's claim?"
format: 
  revealjs:
    slide-number: true
    smaller: true
    chalkboard: false
    scrollable: false
editor: source
---

## Red Marble Proportion Estimation

```{r}
#| echo: true

library(tidyverse)
source("../utils.R")

red_marble = "🔴"
blue_marble = "🔵"

prob_red = 0.64
num_marbles = 5000

set.seed(42)
marbles = sample(c(red_marble, blue_marble), size = num_marbles, 
                 replace = TRUE, prob = c(prob_red, 1 - prob_red))

set.seed(42)
n = 100
one_sample = sample(marbles, n)
p_hat = mean(one_sample == red_marble)
print(str_glue("Percentage of red marbles: {p_hat * 100}%"))
```

- "Half of the marbles in the bag are red!"

## Null and Alternative Hypothesis

:::{.incremental}
- **Null Hypothesis** $H_0$: the parameter is equal to a specific value, $p = 0.5$
- **Alternative Hypothesis** $H_a$/$H_1$: the parameter differs from the value specified by $H_0$
  - Left-tailed: the parameter is <u>less</u> than the value specified by $H_0$, $p < 0.5$
  - Right-tailed: the parameter is <u>greater</u> than the value specified by $H_0$, $p > 0.5$
  - Two-tailed: the parameter is <u>not equal to</u> the value specified by $H_0$, $p \neq 0.5$
:::

## Understanding Hypothesis Testing


:::{.incremental}
- Assume the defendant is innocent.
- The evidence is presented.
- If the evidence strongly indicates guilt, we abandon the presumption of innocence and declare the defendant guilty.
- If not, we maintain the presumption of innocence.
:::


## Assuming Null Hypothesis Is True


```{r}
#| echo: true

p_red = 0.5
n_marbles = 5000

set.seed(42)
m = sample(c(red_marble, blue_marble), size = n_marbles, 
            replace = TRUE, prob = c(p_red, 1 - p_red))

samples = get_samples(m, n, 1000, is_red)
plot_sample_means(samples, n, p_red)

```


## 68-95-99.7 Rule


```{r}
#| echo: true

normal_plot(mean = p_red, sd = sqrt(p_red * (1 - p_red) / n))
```


## Methods

- Critical value method
- $p$-value method

## Significance Level


:::{.incremental}
- The significance level ($\alpha$) is the probability that we use to determine 
whether an event is unusual or statistically significant.
- The most common significance levels are $0.05$ and $0.01$.
- If our test result has a probability less than or equal to this level, 
under the assumption that the null hypothesis is true, 
then the result is sufficiently unusual for us to doubt the null hypothesis.
:::

## Critical Value Method


```{r}
#| echo: false


x = seq(-4, 4, by=0.01)
y = dnorm(x)
df = tibble(x, y)


p1 = ggplot(df, aes(x, y)) +
  geom_line(color="black") +
  stat_function(fun=dnorm, xlim=c(-4,-1.96), fill="red", geom="area") +
  geom_segment(aes(x = -1.96, y = 0, xend = -1.96, yend = dnorm(-1.96)), color = "red") +
  annotate("text", x = -2.2, y = 0.0125, color = "white",
         label = "alpha", size = 3, parse=TRUE) +
  annotate("text", x = -1.96, y = -0.01, color = "black",
         label = "-z[alpha]", size = 3, parse=TRUE) +
  labs(x="", y="", title="Left-tailed Critical Region") +
  theme_minimal() + 
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank())


p2 = ggplot(df, aes(x, y)) +
  geom_line(color="black") +
  stat_function(fun=dnorm, xlim=c(+1.96, +4), fill="red", geom="area") +
  geom_segment(aes(x = +1.96, y = 0, xend = +1.96, yend = dnorm(+1.96)), color = "red") +
  annotate("text", x = 2.2, y = 0.0125, color = "white",
         label = "alpha", size = 3, parse=TRUE) +
  annotate("text", x = 1.96, y = -0.01, color = "black",
         label = "z[alpha]", size = 3, parse=TRUE) +
  labs(x="", y="", title = "Right-tailed Critical Region") +
  theme_minimal() + 
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank())

p3 = ggplot(df, aes(x, y)) +
  geom_line(color="black") +
  stat_function(fun=dnorm, xlim=c(-4,-1.96), fill="red", geom="area") +
  stat_function(fun=dnorm, xlim=c(+1.96, +4), fill="red", geom="area") +
  geom_segment(aes(x = -1.96, y = 0, xend = -1.96, yend = dnorm(-1.96)), color = "red") +
  geom_segment(aes(x = +1.96, y = 0, xend = +1.96, yend = dnorm(+1.96)), color = "red") +
  annotate("text", x = -2.25, y = 0.0125, color = "white",
         label = "alpha/2", size = 3, parse=TRUE) +
  annotate("text", x = 2.25, y = 0.0125, color = "white",
         label = "alpha/2", size = 3, parse=TRUE) +
  annotate("text", x = -1.96, y = -0.01, color = "black",
         label = "-z[alpha/2]", size = 3, parse=TRUE) +
  annotate("text", x = 1.96, y = -0.01, color = "black",
         label = "z[alpha/2]", size = 3, parse=TRUE) +
  labs(x="", y="", title = "Two-tailed Crigical Region") +
  theme_minimal() + 
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank())

grid.arrange(p1, p2, p3, ncol = 3)

```

## Test Statistic Calculation

```{r}
#| echo: true
alpha = 0.05
cv = qnorm(1 - alpha / 2)
print(str_glue("Critical Values are {-round(cv, 3)} and {round(cv, 3)}."))

```

. . .


```{r}
#| echo: true
z_score = (p_hat - p_red) / sqrt(p_red * (1 - p_red) / n)
print(str_glue("z-score: {round(z_score, 3)}"))
```

. . . 

Since $3.8 > 1.96$, we reject the null hypothesis at the 0.05 level 
and conclude that the proportion of red marbles is not 50%.


## $p$-value Method

When using the p-value method in hypothesis testing, we calculate the p-value,
which is the probability of observing a test statistic as extreme as, 
or more extreme than, the one calculated from our sample data, 
assuming that the null hypothesis is true.


## $p$-value Method


```{r}
#| echo: false


x = seq(-4, 4, by=0.01)
y = dnorm(x)
df = tibble(x, y)


p1 = ggplot(df, aes(x, y)) +
  geom_line(color="black") +
  stat_function(fun=dnorm, xlim=c(-4,-2.96), fill="gray39", geom="area") +
  geom_segment(aes(x = -2.96, y = 0, xend = -2.96, yend = dnorm(-2.96))) +
  labs(x="", y="", title="Left-tailed") +
  theme_minimal() + 
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank())


p2 = ggplot(df, aes(x, y)) +
  geom_line(color="black") +
  stat_function(fun=dnorm, xlim=c(+1.46, +4), fill="gray39", geom="area") +
  geom_segment(aes(x = +1.46, y = 0, xend = +1.46, yend = dnorm(+1.46))) +
  labs(x="", y="", title = "Right-tailed") +
  theme_minimal() + 
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank())

p3 = ggplot(df, aes(x, y)) +
  geom_line(color="black") +
  stat_function(fun=dnorm, xlim=c(-4,-2.25), fill="gray39", geom="area") +
  stat_function(fun=dnorm, xlim=c(+2.25, +4), fill="gray39", geom="area") +
  geom_segment(aes(x = -2.25, y = 0, xend = -2.25, yend = dnorm(-2.25))) +
  geom_segment(aes(x = +2.25, y = 0, xend = +2.25, yend = dnorm(+2.25))) +
  labs(x="", y="", title = "Two-tailed") +
  theme_minimal() + 
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank())

grid.arrange(p1, p2, p3, ncol = 3)

```

## $p$-value Calculation

$$X \sim N(0.5, 0.0025)$$

. . .

$$Prob(x >= 0.69) = ?$$

. . .

```{r}
#| echo: true
p_val = (1 - pnorm(p_hat, mean = 0.5, sd = 0.05)) * 2
print(str_glue("p-value: {p_val}"))

```

. . .

Since $0.00014 < 0.05$, we reject the null hypothesis at the 0.05 level 
and conclude that the proportion of red marbles is not 50%.

## Making Mistakes
    
```{r}
#| echo: true

set.seed(7595)
n = 100
one_sample = sample(marbles, n)
p_hat = mean(one_sample == red_marble)
print(str_glue("Percentage of red marbles: {p_hat * 100}%"))
```

. . .

```{r}
#| echo: true

z_score = (p_hat - p_red) / sqrt(p_red * (1 - p_red) / n)
print(str_glue("z-score: {round(z_score, 3)}"))

p_val = (1 - pnorm(p_hat, mean = 0.5, sd = 0.05)) * 2
print(str_glue("p-value: {round(p_val, 3)}"))

```

. . . 

Since $-1.96 < 1.6 < 1.96$ or $0.11 > 0.5$, we fail to reject the null hypothesis at the 0.05 level 
and conclude that the proportion of red marbles is 50%.

. . . 

**Type II error**: fail to reject the null hypothesis when it is false.

## Type I and II Errors

| | $H_0$ is true | $H_0$ is false |
| --- | --- | --- |
| Reject $H_0$ | Type I error | Correct decision |
| Fail to reject $H_0$ | Correct decision | Type II error |


## Interview Questions

[**Facebook**: Say you flip a coin 10 times and observe only one head.
What would be your null hypothesis and $p$-value for testing whether
the coin is fair or not.]{style="color: blue"}

. . . 

Let $p$ be the probability of getting a head.

$H_0$: $p = 0.5$; $H_a$: $p \neq 0.5$

. . .

[CLT does not apply because the sample size is less than 30.]{style="color: red"}

. . . 

$p$-value: probability of observing an outcome as extreme as or more extreme than
the observed data, assuming the null hypothesis is true.

. . .

$X\sim \text{Binomial}(n=10, p=0.5)$

$$Porb(x=0) + Prob(x = 1) + Prob(x=9) + Prob(x=10)$$

. . . 

```{r}
#| echo: true
p_val = (choose(10, 1) * 0.5^1 * 0.5^9 + choose(10, 0) * 0.5^0 * 0.5^10) * 2
print(str_glue("p-value: {p_val}"))
```

. . .

Since `r round(p_val, 3)` < 0.05, we reject the null hypothesis at the 0.05 level
and conclude that the coin is not fair.



## Interview Questions

[**D.E. Shaw**: A coin was flipped 1,000 times, and 550 times it showed heads. Do you
think the coin is biased? Why or why not?]{style="color: blue"}

. . .


Let $p$ be the probability of getting a head.

$H_0$: $p = 0.5$; $H_a$: $p \neq 0.5$

. . .


```{r}
#| echo: true

z_score = (0.55 - 0.5) / sqrt(0.5 * (1 - 0.5) / 1000)
print(str_glue("z-score: {round(z_score, 3)}"))


p_val = (1 - pnorm(0.55, 0.5, sqrt(0.5 * (1 - 0.5) / 1000)))
print(str_glue("p-value: {round(p_val, 3)}"))
```

. . .

Because `r z_score` > 1.96 and `r p_val` < 0.05, we reject the null hypothesis at 
the 0.05 level and conclude that the coin is not fair.












